Tail Flexibility in the Degrees of Preferential Attachment Networks

### 2025-06-23 paper.tex
Updating paper draft

\abstract{Devising the underlying generating mechanism of a real-life
network is difficult as, more often than not, only its snapshots are
available, but not its full evolution. One candidate for the generating
mechanism is preferential attachment which, in its simplest form,
results in a degree distribution that follows the power law.
Consequently, the growth of real-life networks that roughly display such
power-law behaviour is commonly modelled by preferential attachment.
However, the validity of the power law has been challenged by the
presence of alternatives with comparable performance, as well as the
recent findings that the right tail of the degree distribution is often
lighter than implied by the body, whilst still being heavy. In this
paper, we study a modified version of the model with a flexible
preference function that allows super/sub-linear behaviour whilst also
guaranteeing that the limiting degree distribution has a heavy tail. We
relate the distributions tail heaviness directly to the model
parameters, allowing direct inference of the parameters from the degree
distribution alone.}

@@

Networks appear in many fields such as sociology, politics,
epidemiology, and economics. Statistical methods have been and continue
to be used to study networks that appear in these fields and beyond,
from the use of stochastic block models to detect communities within the
French political blogosphere \citep{Latouche11}, to using Exponential
Random Graph Models (ERGMs) to analyse the global trade network
\citep{Setayesh22}, and the use of mechanistic models in neuroscience to
investigate the patterns of connections in neural systems
\citep{Betzel17}.

One example of a mechanistic model that may be considered is the
preferential attachment (PA) model where as a network grows and new
vertices join they connect to those already present in the network with
weights proportional to some preference function \(b(\cdot)\) of their
in-degree. The well-known Barabási-Albert (BA) model is a special case
of this when \(b(k) = k+1\) \citep{Barabasi99}.

These mechanistic models are generally useful when simulating networks,
but not for fitting to real data as usually only snapshots of a network
are available and not the full evolution - without this conclusions
about the growth are difficult to make. However, one statistic that is
usually available for snapshots of networks is the degree distribution
making it valuable to study the degrees in a network. Often, when
looking at the degree distributions of real networks it can be
attractive to use the power law \footnote{Power law:
  \(\bar F(k) = L(k)\cdot k^{-1/\xi}\), where \(\xi>0\) and \(L(k)\) is
  slowly varying in \(k\).} when modelling as it is seemingly observed
in much of the data. It also comes with the added benefit of offering a
connection to preferential attachment since the BA model results in a
power law degree distribution \citep{Barabasi99}, perhaps suggesting
that if the power law fits well then the network may have grown
according to similar simple rules. However, attempting to explain the
networks growth using preferential attachment when the degrees seemingly
follow a power law may not be the best course of action as
\citep{krapivsky01} shows that a general preference function does not
necessarily lead to power law degrees, showing that a preference
function of form \(b(k) = k^\alpha\) leads to light tails when
\(\alpha<1\), a degenerate network (a finite number of vertices end up
with all of the added edges) when \(\alpha>1\), and only yielding heavy
tails when \(\alpha=1\). Additionally, it is still being argued whether
or not a lot of real data actually follows a power law.
\citet{Broido_2019} provides some evidence that suggests that real world
networks rarely have degrees that follow a power law through the use of
statistical analysis of almost 1000 networks across a wide variety of
fields; \citet{Voitalov_2019}, on the other hand, disagrees and claims
that these kinds of networks are not nearly as rare and only appear so
as a result of an unrealistic expectation of power law without
deviations or noise. \citet{rudas07} derives the limiting degree
distribution in terms of the preference function \(b(\cdot)\) and its
parameters, we build upon this by finding a class of preference
functions that result in realistic tails i.e.~heavy but not as heavy as
implied by the power law.

Extreme value methods have been scarcely used when it comes to analysis
of degree distributions and whether they follow the power law. The
absence of these methods essentially down weights the deviation of
vertices with the largest degrees, which are usually the most
influential, from the power law. Somewhat recent publications
\citep{Voitalov_2019, Lee24, wang2022random} have addressed this gap in
the literature. \citet{Voitalov_2019} proposes estimating the power law
exponent of a degree distribution using three estimators based in
extreme value theory, in contrast to the estimators generally used in
network science, the degree distributions are then split into one of
three categories (not power-law , hardly power-law or power-law)
depending on the values of the estimates obtained from each of the
estimators. \citet{Lee24} proposes a framework for modelling entire
degree distributions using a combination of two generalisations of the
power law including a model selection step that is capable of suggesting
whether or not the data is adequately modelled by a power law, with one
conclusion being that most data seems to have a tail lighter than what
is implied by the body whilst still being heavy tailed.
\citet{wang2022random} considers a preferential attachment model with an
affine preference function and reciprocal edges, they study the limiting
degree distribution of this model for both the in-degrees and
out-degrees - finding that the joint limiting distribution is
multivariate regularly varying and has the property of hidden regular
variation.

Evidently, the study of degree distributions is a very active but the
vast majority of research in the field (whether using extremes or not)
are purely descriptive in the sense that no information about the
preference function is revealed even if preferential attachment was the
underlying mechanism. This paper aims to address this gap - asking if
given the degree distribution of a network assumed to come from the
preferential attachment model, can we directly infer the model
parameters?

The remainder of this paper begins with a more detailed description of
the PA model alongside the theoretical results for the limiting degree
distribution using results from \citet{rudas07},
(Section~\ref{sec-tail}) then goes on to analyse the tail behaviour of
this limiting distribution in terms of the preference function
\(b(\cdot)\) arguing that a heavy tailed degree distribution can only be
the result of an asymptotically linear preference function shown by
Proposition~\ref{prp-omega} - ending with a suggestion for a preference
function that can guarantee heavy tails while remaining flexible.
Section~\ref{sec-model} builds on the results from
Section~\ref{sec-tail} and considers the limiting distribution obtained
when using a preference function of the form stated at the end of
Section~\ref{sec-tail} providing examples of how the shape of the
distribution varies with respect to the model parameters as well as
visualisations for how the tail heaviness changes when these parameters
are varied - continued study of this distribution then also reveals a
connection to the general Pareto distribution (GPD).
Section~\ref{sec-sim} consists of a simulation study where networks are
simulated from the PA model for various parameter combinations and then
the model specified in Section~\ref{sec-model} is fitted to the
resulting degree distributions using Bayesian inference, demonstrating
that (for simulated data) the model parameters can be recovered from
only the degree distribution. Finally, Section~\ref{sec-real} fits the
model to some real data and provides posterior estimates for the
preference function assuming the networks evolved according to the PA
model.

@@

The network generative model that we will be focussing on in this paper
is dubbed General Preferential Attachment in \citep{rudas07} and is
defined as follows:

Starting at time \(t=0\) with an initial network of \(m\) vertices that
each have no edges, at times \(t=1,2,\ldots\) a new vertex is added to
the network bringing with it \(m\) directed edges (with the new vertex
as the source); the target for each of these edges are selected from the
vertices already in the network with weights proportional to some
function \(b\) of their degree where the preference function \(b\) is
chosen such that:

\begin{equation}\phantomsection\label{eq-condb1}{
b:\mathbb N \mapsto \mathbb R^+\setminus\{0\},
}\end{equation}

@@

Special cases of this model include the Barabási-Albert (BA) model when
\(b(k) = k+\varepsilon\), which leads to a power-law degree distribution
with \(\xi=1/2\) and the Uniform Attachment (UA) model where \(b(k)=c\)
leading to a degree distribution that is light tailed.

Given conditions \ref{eq-condb1} and \ref{eq-condb2} an expression for
the survival of the limiting degree distribution can be found in the
case that \(m=1\); obtained by considering a branching process that is
equivalent to the growth of the network, as in \citep{rudas07}. Theorem
1 from \citep{rudas07} states that for the tree \(\Upsilon(t)\) at time
\(t\):

\begin{equation}\phantomsection\label{eq-survlim}{
\lim_{t\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum_{x\in\Upsilon(t)}\varphi(\Upsilon(t)_{\downarrow x}) = \lambda^* \int_0^\infty e^{-\lambda^* t}\mathbb E\left[\varphi(\Upsilon(t))\right]dt
}\end{equation} where \(\lambda^*\) satisfies \(\hat\rho(\lambda^*)=1\).

The limiting survival can be viewed as the limit of the empirical
proportion of vertices with degree over a threshold \(k\), that is:

@@

\[
\bar F(k) = \frac{\int_0^\infty e^{-\lambda^* t}\mathbb E\left[\mathbb I\left\{\text{deg}(x,\Upsilon(t))>k\right\}\right]dt}{\int_0^\infty e^{-\lambda^* t}dt} = \prod_{i=0}^k\frac{b(i)}{\lambda^* + b(i)}
\] Additionally, using the fact that \(f(k) = \bar F(k-1) - \bar F(k)\)
the limiting degree distribution can be shown to be:

\[
f(k) = \frac{\lambda^*}{\lambda^* + b(k)}\prod_{i=0}^{k-1}\frac{b(i)}{\lambda^*+b(i)},\qquad k\in\mathbb N.
\]

We are interested in how the tail behaviour of the limiting degrees is
affected by the preference function \(b\).

\citep{shimura12} introduces a quantity that will help in determining
this for this discrete distribution, that is:

For a distribution \(F\) with survival function \(\bar F\) and some
\(k\in\mathbb Z^+\) let:

\[
\Omega(F,k) = \left(\log\displaystyle\frac{\bar F (k+1)}{\bar F (k+2)}\right)^{-1} - \left(\log\displaystyle\frac{\bar F (k)}{\bar F (k+1)}\right)^{-1}
\]

\citep{shimura12} then states that if
\(\lim_{n\rightarrow\infty} \Omega(F,k) = 1/\alpha\) (\(\alpha>0\)),
then \(F\) is heavy tailed with \(\bar F(k) \sim k^{-\alpha}\).
Additionally, if \(\lim_{n\rightarrow\infty} \Omega(F,k) = 0\) then the
distribution is light tailed. This allows us to show the following:

@@

If \(b(k) \rightarrow \infty\) as \(k\rightarrow \infty\) then \[
\bar F(k) = \prod_{i=0}^k\frac{b(i)}{\lambda^* + b(i)}
\] and

\[
\lim_{k\rightarrow\infty}\Omega(F,k) = \lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{\lambda^*}.
\]

@@

See appendix \ref{sec-proofs} for the details of the proof.

Proposition~\ref{prp-omega} aligns with the result from
\citep{krapivsky01} demonstrating that a sub-linear preference function
will lead to a light tailed distribution. So, in order for the degree
distribution to be heavy tailed we need that the limit
\(\lim_{k\rightarrow\infty} b(k+1)-b(k)\) exists and is positive. We can
in fact show that BA model produces a heavy tailed degree distribution
with index \(\xi=0.5\) by considering the preference function
\(b(k) = k + \varepsilon\), noting that the limit
\(\lim_{k\rightarrow\infty} b(k+1)-b(k)\) is simply
\(\lim_{k\rightarrow\infty} 1 = 1\) leaving the tail heaviness to be
\(1/\lambda^*\) which using \(\hat\rho\) can be found to be \(1/2\).

@@

Let \((a_n)_{n\ge1}\) and \((b_n)_{n\ge1}\) be two sequence of real
numbers. Assume \((a_n)_{n\ge1}\) is a strictly monotone and divergent
sequence and the following limit exists:

@@

Consider a GPA model with preference function \(b(\cdot)\) satisfying
Proposition~\ref{prp-omega}. If
\(\lim_{k\rightarrow\infty}\frac{b(k)}{k} = c\), then the limiting
degree distribution has a heavy tail with index \(c\over\lambda^*\) and
this heavy tail may only be the result of a preference function of this
form (asymptotically linear).

@@

\[
b(k) = \begin{cases}
k^\alpha + \varepsilon,&k<k_0\\
k_0^\alpha + \varepsilon + \beta(k-k_0), &k\ge k_0
\end{cases}
\] for \(\alpha,\beta, \varepsilon>0\) and \(k_0\in\mathbb N\).

This preference function, as per Proposition~\ref{prp-omega}, will
produce a degree distribution with tail heaviness \(\beta/\lambda^*\)
guaranteeing a heavy tail. We study this preference function further in
the next section.

@@

Using a preference function with guaranteed linear behaviour in the
limit, allows for the inclusion of sub/super linear behaviour without
losing the heavy tails or ending up with a degenerate degree
distribution.

The limiting degree distribution resulting from using a preference
function of this form can be found to have survival:

\begin{equation}\phantomsection\label{eq-polysurv}{
\bar F(k) = \begin{cases}
\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0\\
\left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right)\frac{\Gamma(\lambda^*+k_0^\alpha + \varepsilon)/\beta)}{\Gamma\left((k_0^\alpha + \varepsilon)/\beta\right)} \frac{\Gamma\left(k-k_0 + 1 +\frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0 + 1 +\frac{\lambda^* +k_0^\alpha + \varepsilon}{\beta}\right)},&k\ge k_0.
\end{cases}
}\end{equation}

with \(\lambda^*\) satisfying \(\hat \rho(\lambda^*)=1\) where:

@@

Some examples of what the degree distribution looks like for various
parameters are shown below in Figure~\ref{fig-polylinsurv}:

@@

The survival function (\ref{eq-polysurv}) can be connected to a
discretised variation of the generalised Pareto distribution (GPD)
usually dubbed the Integer GPD (IGPD) with survival: \[
\Pr(X> x|X> v) = \left(\frac{\xi(x-v)}{\sigma} + 1\right)^{-1/\xi},\qquad x=v+1,v+2,\ldots
\] for \(v\in\mathbb Z^+, \sigma>0,\xi\in \mathbb R\), denoted as
\(X|X>u \sim  \mathrm {IGPD}(\xi, \sigma, u)\) where \(\xi\) is the
shape parameter controlling th tail heaviness.

@@

\[
\bar F(k) 
\begin{cases}
=\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0,\\
\approx \left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right) \left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0,
\end{cases}
\] meaning that for \(k\ge k_0\) the limiting degree distribution (for
large \(k_0^\alpha\)) is approximated by
\(\text{IGPD}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon}{\lambda^*},k_0-1\right)\).

To assess how close of an approximation this is, the theoretical
conditional survivals are shown in Figure~\ref{fig-approx_surv} in
colour and their IGPD approximations are shown in grey. The
approximation seems to hold up fairly well even for large degrees.

@@

\caption{\label{fig-approx_surv}Theoretical conditional survivals (grey)
alongside their IGPD approximations (coloured).}

@@

In agreement with Proposition~\ref{prp-omega2}, \(\beta>0\), the shape
parameter of the IGPD is positive and thus the distribution is heavy
tailed. Additionally the value of the shape parameter \(\xi\) is shown
in Figure~\ref{fig-polyheat} for various parameter choices. The darker
regions on the heat maps correspond to a heavier tail and the lighter to
a lighter tail, the red dashed line shows combinations of \(\alpha\) and
\(\beta\) that produce a limiting degree distribution with the same tail
heaviness as the Barabási-Albert model, \(\xi=0.5\).

@@

Finally to be able to perform inference of the model parameters, we
consider a network with degree count vector
\(\pmb n = (n_0, n_1, \ldots, n_M)\) and maximum degree \(M\). We can
then write the likelihood of this model for this network as:

\begin{align*}
L(\pmb x,\pmb n | \pmb \theta) = \left(\frac{\lambda^*}{\lambda^*+\varepsilon}\right)^{n_0}\left(\prod_{j=l}^{k_0-1}\frac{j^\alpha +\varepsilon}{\lambda^* + j^\alpha +\varepsilon}\right)^{\left(\sum_{i\ge k_0}n_{i}\right)} \prod_{l \le i<k_0}\left(\frac{\lambda^*}{\lambda^* +i^\alpha + \varepsilon } \prod_{j=l}^{k_0-1}\frac{j^\alpha + \varepsilon}{\lambda^* + j^\alpha + \varepsilon}\right)^{n_i}\\ \times \prod_{i\ge k_0}\left(\frac{\text{B}(i-k_0 + (k_0^\alpha + \varepsilon)/\beta,1+\lambda^*/\beta)}{\text{B}((k_0^\alpha + \varepsilon)/\beta,\lambda^*/\beta)}\right)^{n_i}
\end{align*}

where \(B(y,z)\) is the the beta function and \(l\ge0\) is a variable
that allows truncating the data such that the minimum degree is \(l\),
which will allow the model to be fitted whilst ignoring the influence of
the lower degrees (those less than \(l\)) as the model does not capture
the behaviour at the lower degrees since \citet{rudas07} only provides
results for the case of a preferential attachment tree.

\section{Simulation Study}\label{sec-sim}

This section aims to show that the parameters of the model in
Section~\ref{sec-model} can be recovered from the degree distribution of
a network simulated from it.

@@

to obtain a posterior distribution that can then be used in an adaptive
Metropolis-Hastings Markov chain Monte Carlo (MCMC) algorithm to obtain
posterior samples. The results of this inference are shown in
Figure~\ref{fig-rec1} and Figure~\ref{fig-rec2}. For these simulated
networks \(l=0\).

@@

{\centering \pandocbounded{\includegraphics[keepaspectratio]{images/pars_plot.png}}

>>>>

\abstract{Devising the underlying generating mechanism of a real-life
network is difficult as, more often than not, only its snapshots are
available, but not its full evolution. One candidate for the generating
mechanism is preferential attachment which, in its simplest form,
results in a degree distribution that follows the power law.
Consequently, the growth of real-life networks that roughly display such
power-law behaviour is commonly modelled by preferential attachment.
However, the validity of the power law has been challenged by the
presence of alternatives with comparable performance, as well as the
recent findings that the right tail of the degree distribution is often
lighter than implied by the body, whilst still being heavy. In this
paper, we study a modified version of the model with a flexible
preference function that allows super/sub-linear behaviour whilst also
guaranteeing that the limiting degree distribution has a heavy tail. We
relate the distributions tail index directly to the model parameters,
allowing direct inference of the parameters from the degree distribution
alone.}

@@

Networks appear in many fields such as sociology, politics,
epidemiology, and economics. Statistical methods have been used to study
networks that appear in these fields and beyond, from the use of
stochastic block models to detect communities within the French
political blogosphere \citep{Latouche11}, to using Exponential Random
Graph Models (ERGMs) to analyse the global trade network
\citep{Setayesh22}, and the use of mechanistic models in neuroscience to
investigate the patterns of connections in neural systems
\citep{Betzel17}.

One example of mechanistic models is the preferential attachment (PA)
model, where as a network grows and new vertices join, they connect to
those already present in the network with probabilities proportional to
some preference function of their in-degrees. The well-known
Barabási-Albert (BA) model is a special case of this when \(b(k) = k+1\)
\citep{Barabasi99}.

These mechanistic models are generally useful when simulating networks,
but not for fitting to real data as usually only snapshots of a network
are available and not the full evolution - without this conclusions
about the growth are difficult to make. However, one statistic that is
usually available for snapshots of networks is the degree distribution,
making it valuable to study the degrees in a network. Often, it is
appealing to model the degree distributions of real networks using the
power law \footnote{Power law: \(\bar F(k) = L(k)\times k^{-1/\xi}\),
  where \(\xi>0\) and \(L(k)\) is slowly varying in \(k\).}, as it is
seemingly observed in much of the data. Furthermore, as the BA model
results in a power law degree distribution \citep{Barabasi99}, that the
power law fits well offers the interpretation that the network may have
grown according to PA or similar simple rules. However, attempting to
explain the networks growth using preferential attachment when the
degrees seemingly follow a power law may not be the best course of
action as \citet{krapivsky01} show that a general preference function
does not necessarily lead to power law degrees. Specifically, showing
that a preference function of form \(b(k) = k^\alpha\) leads to light
tails when \(\alpha<1\), a degenerate network (a finite number of
vertices end up with all of the added edges) when \(\alpha>1\), and only
yields heavy tails when \(\alpha=1\). \citet{rudas07} derives the
limiting degree distribution in terms of the preference function
\(b(\cdot)\) and its parameters, we build upon this by finding a class
of preference functions that result in realistic tails i.e.~heavy but
not as heavy as implied by the power law. On the other hand, it is still
being argued whether or not many real data actually follow the power
law. \citet{Broido_2019} provides some evidence that real world networks
rarely have degrees that follow a power law, through the use of
statistical analysis of almost 1000 networks across a wide variety of
fields; \citet{Voitalov_2019}, on the other hand, disagrees and claims
that these kinds of networks are not nearly as rare and only appear so
as a result of an unrealistic expectation of power law without
deviations or noise.

Extreme value methods have been scarcely used when it comes to analysis
of degree distributions and whether they follow the power law. The
absence of these methods essentially downweights the deviation of
vertices with the largest degrees, which are usually the most
influential, from the power law. \citet{Voitalov_2019} proposes
estimating the power law exponent of a degree distribution using three
estimators based in extreme value theory, in contrast to the estimators
generally used in network science, the degree distributions are then
split into one of three categories (not power-law, hardly power-law or
power-law) depending on the values of the estimates obtained from each
of the estimators. \citet{Lee24} proposes a framework for modelling
entire degree distributions using a combination of two generalisations
of the power law with a model selection step that is capable of
suggesting whether or not the data is adequately modelled by a power
law, with one conclusion being that most data seems to have a tail
lighter than what is implied by the power law in body whilst still being
heavy tailed. One of the models used by \citet{Lee24} is the Zipf-IGP
model where up until a certain threshold the data is modelled using the
Zipf disribution (a discrete power law) and beyond it a discrete version
of the Generalised Pareto distribution is used, we will later compare
our model to this as they share some similarities.
\citet{wang2022random} considers a preferential attachment model with an
affine preference function and reciprocal edges, finding that the joint
limiting distribution of the in-degrees and out-degrees is multivariate
regularly varying and has the property of hidden regular variation.

Evidently, the study of degree distributions is a very active but the
vast majority of research in the field (whether using extremes or not)
are purely descriptive in the sense that no information about the
preference function is revealed even if preferential attachment was the
underlying mechanism. This paper aims to address this gap, asking if the
degree distribution of a network is assumed to come from the
preferential attachment model, can we directly infer the model
parameters?

The remainder of this paper is as follows: Section 2 gives a detailed
description of the PA model alongside the theoretical results for the
limiting degree distribution with a focus on its tail behaviour in terms
of the preference function \(b(\cdot)\). It is shown that a heavy tailed
degree distribution can only be the result of an asymptotically linear
preference function ending with a suggestion for a preference function
that can guarantee heavy tails while remaining flexible.
Section~\ref{sec-model} utilises the preference function proposed at the
end of Section 2, and illustrates how the tail index of the degree
distribution varies with the model parameters. Section~\ref{sec-sim}
consists of a simulation study where networks are simulated from the
proposed model for various parameter combinations, demonstrating that
the parameters can be recovered from fitting the model to only the
degree distribution. Section~\ref{sec-real} fits the model to some real
data and provides posterior estimates for the preference function.
Section~\ref{sec-conc} concludes the article.

@@

The network generative model that we will focus on in this paper is the
General Preferential Attachment model in \citet{rudas07} and is defined
as follows:

Starting at time \(t=0\) with an initial network of \(m\) vertices that
each have no edges, at times \(t=1,2,\ldots\) a new vertex is added to
the network bringing with it \(m\) directed edges (with the new vertex
as the source); the target for each of these edges are selected from the
vertices already in the network with weights proportional to some
preference function \(b(\cdot)\) of their degree, where
\(b: N \mapsto \mathbb R^+\setminus\{0\}\) is such that:

@@

Special cases of this model include the Barabási-Albert (BA) model when
\(b(k) = k+\varepsilon\), which leads to a power-law degree distribution
with \(\xi=1/2\), and the Uniform Attachment (UA) model where \(b(k)=c\)
leading to a degree distribution that is light tailed.

Given condition \ref{eq-condb2} the survival function of the limiting
degree distribution, called the limiting survival hereafter, can be
analytically derived in the case where \(m=1\), obtained by considering
a branching process that is equivalent to the growth of the network, as
in \citet{rudas07}. Theorem 1 from \citet{rudas07} states that for the
tree \(\Upsilon(t)\) at time \(t\):

\begin{equation}\phantomsection\label{eq-survlim}{
\lim_{t\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum_{x\in\Upsilon(t)}\varphi(\Upsilon(t)_{\downarrow x}) = \lambda^* \int_0^\infty e^{-\lambda^* t}\mathbb E\left[\varphi(\Upsilon(t))\right]dt
}\end{equation} where \(\lambda^*\) satisfies \(\hat\rho(\lambda^*)=1\)
and \(\hat\rho\) is the Laplace transform of the density of the point
process associated with the pure birth process that corresponds to a
nodes individual growth, that is:

\[
\hat\rho(\lambda) \coloneq \int_0^\infty e^{-\lambda t}\rho(t)\mathrm{d}t.
\]

The limiting survival can be viewed as the limit of the empirical
proportion of vertices with degree over a threshold \(k\in\mathbb N\),
that is:

@@

\[
\bar F(k) = \frac{\int_0^\infty e^{-\lambda^* t}\mathbb E\left[\mathbb I\left\{\text{deg}(x,\Upsilon(t))>k\right\}\right]dt}{\int_0^\infty e^{-\lambda^* t}dt} = \prod_{i=0}^k\frac{b(i)}{\lambda^* + b(i)}.
\] Therefore, the corresponding probability mass function of the degree
distribution \(f(k) = \bar F(k-1) - \bar F(k)\) is

\[
f(k) = \frac{\lambda^*}{\lambda^* + b(k)}\prod_{i=0}^{k-1}\frac{b(i)}{\lambda^*+b(i)}.
\]

We are interested in how the tail behaviour of the discrete limiting
degree distribution is affected by the preference function \(b\), which
can be determined by the results of \citet{shimura12}.

For a distribution \(F\) with survival function \(\bar F\) and some
\(k\in\mathbb Z^+\) let

\[
\Omega(F,k) = \left(\log\displaystyle\frac{\bar F (k+1)}{\bar F (k+2)}\right)^{-1} - \left(\log\displaystyle\frac{\bar F (k)}{\bar F (k+1)}\right)^{-1}.
\]

\citet{shimura12} states that if
\(\lim_{k\rightarrow\infty} \Omega(F,k) = 1/\alpha\) (\(\alpha>0\)),
then \(F\) is heavy-tailed with \(\bar F(k) \sim k^{-\alpha}\). On the
other hand, if \(\lim_{k\rightarrow\infty} \Omega(F,k) = 0\) then the
distribution is light-tailed. This allows us to show the following:

@@

If \(\bar F(k) = \prod_{i=0}^k\frac{b(i)}{\lambda^* + b(i)}\) and
\(b(k) \rightarrow \infty\) as \(k\rightarrow \infty\), then \[
\lim_{k\rightarrow\infty}\Omega(F,k) = \lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{\lambda^*}.
\]

@@

See Appendix \ref{sec-proofs} for the details of the proof.

Proposition~\ref{prp-omega} aligns with the result from
\citet{krapivsky01} demonstrating that a sub-linear preference function
will lead to a light-tailed distribution, as
\(\lim_{k\rightarrow\infty} b(k+1)-b(k) = 0\) if \(b(k)=k^\alpha\) where
\(\alpha < 1\). Proposition 2.1 also aligns with the fact that BA model
produces a heavy-tailed degree distribution with index \(\xi=0.5\) by
considering the preference function \(b(k) = k + \varepsilon\), as
\(\lim_{k\rightarrow\infty}b(k+1)-b(k)=1\) leaving the tail index to be
\(1/\lambda^*\) which using \(\hat\rho\) can be found to be \(1/2\). So,
in order for the degree distribution to be heavy-tailed we need that the
limit \(\lim_{k\rightarrow\infty} b(k+1)-b(k)\) exists and is positive.
To determine the class of functions that will result in heavy-tailed
limiting degree distributions, we use the following result:

@@

Let \((a_k)_{k\ge1}\) and \((b_k)_{k\ge1}\) be two sequence of real
numbers. Assume \((a_k)_{k\ge1}\) is a strictly monotone and divergent
sequence and the following limit exists:

@@

Consider a GPA model with preference function \(b(\cdot)\) satisfying
Proposition~\ref{prp-omega}. Then the limiting degree distribution has a
heavy tail with index \(c/\lambda^{*}\) if and only if
\(\lim_{k\rightarrow\infty}b(k)/k = c > 0\).

@@

\begin{equation}\phantomsection\label{eq-pref}{
b(k) = \begin{cases}
k^\alpha + \varepsilon,&k<k_0,\\
k_0^\alpha + \varepsilon + \beta(k-k_0), &k\ge k_0
\end{cases}
}\end{equation} for \(\alpha,\beta, \varepsilon>0\) and
\(k_0\in\mathbb N\).

This preference function, as per Proposition~\ref{prp-omega}, will
produce a degree distribution with tail index \(\beta/\lambda^*\)
guaranteeing a heavy tail. We study this preference function further in
the next section.

@@

In the previous section, we found that using a preference function with
linearity in the limit allows for the inclusion of sub/super-linear
behaviour below the threshold, while simultaneously guaranteeing a heavy
tail. In this section, we study how the limiting degree distribution
varies with the preference function according to Equation~\ref{eq-pref}.
Its survival is

\begin{equation}\phantomsection\label{eq-polysurv}{
\bar F(k) = \begin{cases}
\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0,\\
\left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right)\frac{\Gamma(\lambda^*+k_0^\alpha + \varepsilon)/\beta)}{\Gamma\left((k_0^\alpha + \varepsilon)/\beta\right)} \frac{\Gamma\left(k-k_0 + 1 +\frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0 + 1 +\frac{\lambda^* +k_0^\alpha + \varepsilon}{\beta}\right)},&k\ge k_0,
\end{cases}
}\end{equation}

with \(\lambda^*\) satisfying \(\hat \rho(\lambda^*)=1\) where

@@

Some examples of the limiting degree distribution for various parameter
combinations are shown below on log-log scale in
Figure~\ref{fig-polylinsurv}:

\begin{verbatim}
     a   b  eps k0 x   lambda        surv
10 0.5 1.5 0.01 20 0 1.500094 0.006622105
\end{verbatim}

\begin{verbatim}
[1] 0.9999374
\end{verbatim}

@@

Figure~\ref{fig-polylinsurv} demonstrates that this model can capture a
wide variety of shapes for the survival function, including a large
range of possible tail indices ranging from 0.035
(\(\alpha=1.5, \beta=0.1, \varepsilon=1\)) to 0.999
(\(\alpha=0.5, \beta=1.5, \varepsilon=0.01\)).

The survival function (\ref{eq-polysurv}) can be connected to the
discrete version of the generalised Pareto distribution (GP), called the
Integer GP (IGP) \citep{Rohrbeck_2018} with conditional survival: \[
\Pr(X> x|X> v) = \left(\frac{\xi(x-v)}{\sigma} + 1\right)^{-1/\xi},\qquad x=v+1,v+2,\ldots
\] for \(v\in\mathbb Z^+, \sigma>0,\xi\in \mathbb R\), denoted as
\(X|X>u \sim  \mathrm {IGP}(\xi, \sigma, u)\) where \(\xi\) is the shape
parameter controlling the tail index

@@

Therefore, \[
\bar F(k) 
\begin{cases}
=\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0,\\
\approx \left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right) \left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0,
\end{cases}
\] meaning that for \(k\ge k_0\) the limiting degree distribution (for
large \(k_0^\alpha\)) is approximated by
\(\text{IGP}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon}{\lambda^*},k_0-1\right)\).

To assess how close of an approximation this is, the theoretical
conditional survivals are shown in Figure~\ref{fig-approx_surv} in
colour and their IGP approximations are shown in grey. The approximation
seems to hold up fairly well even for large degrees and more so when
\(\alpha\) is larger.

@@

\caption{\label{fig-approx_surv}Theoretical conditional survivals (grey)
alongside their IGP approximations (coloured).}

@@

In agreement with Proposition~\ref{prp-omega2}, \(\beta>0\) ensures that
the shape parameter of the IGP is positive and thus the distribution is
heavy-tailed. Additionally the value of the shape parameter \(\xi\) is
shown in Figure~\ref{fig-polyheat} for various parameter choices. The
darker and lighter regions on the heat maps correspond to a heavier and
a lighter tail, respectively, the red dashed line shows combinations of
\(\alpha\) and \(\beta\) that produce a limiting degree distribution
with the same tail index as the Barabási-Albert model, that is,
\(\xi=0.5\).

@@

To perform inference of the model parameters, we consider a network with
degree count vector \(\pmb n = (n_0, n_1, \ldots, n_M)\), where \(M\) is
the maximum degree. We can then write the likelihood as:

\begin{align*}
L(\pmb x,\pmb n | \pmb \theta) = &\left(\frac{\lambda^*}{\lambda^*+\varepsilon}\right)^{n_0}\left(\prod_{j=l}^{k_0-1}\frac{j^\alpha +\varepsilon}{\lambda^* + j^\alpha +\varepsilon}\right)^{\left(\sum_{i\ge k_0}n_{i}\right)} \\ &\times \prod_{l \le i<k_0}\left(\frac{\lambda^*}{\lambda^* +i^\alpha + \varepsilon } \prod_{j=l}^{k_0-1}\frac{j^\alpha + \varepsilon}{\lambda^* + j^\alpha + \varepsilon}\right)^{n_i}\\ &\times \prod_{i\ge k_0}\left(\frac{\text{B}(i-k_0 + (k_0^\alpha + \varepsilon)/\beta,1+\lambda^*/\beta)}{\text{B}((k_0^\alpha + \varepsilon)/\beta,\lambda^*/\beta)}\right)^{n_i}
\end{align*}\label{eq-lh}

where \(B(y,z)\) is the the beta function, and \(l\ge0\) is a variable
that allows truncating the data such that the minimum degree is \(l\).
This will allow the model to be fitted whilst ignoring the influence of
the lower degrees (those less than \(l\)) as the model does not capture
the behaviour well at the lower degrees since \citet{rudas07} only
provides results for the case of a preferential attachment tree.

\section{Applications}\label{applications}

\subsection{Simulated Data}\label{sec-sim}

This first subsection aims to show that the parameters of the model in
Section~\ref{sec-model} can be recovered from simulating a network from
the model, and fitting it to the observed degree distribution, using the
likelihood in \eqref{eq-lh}.

@@

where Ga(a, b) is the Gamma distribution with shape a and rate b, and
U(a, b) is uniform distribution with lower and upper bounds a and b, to
obtain a posterior distribution, up to the proportionality constant.
Posterior samples can then be obtained by an adaptive
Metropolis-Hastings Markov chain Monte Carlo (MCMC) algorithm. For these
simulated networks \(l=0\).

@@

\section{Conclusion and Discussion}\label{sec-conc}
